
% ZEO
%    Installing ZEO
%    How ZEO works (ClientStorage)
%    Configuring ZEO
   
\section{ZEO}
\label{zeo}

\subsection{How ZEO Works}

The ZODB, as I've described it so far, can only be used within a
single Python process (though perhaps with multiple threads).  ZEO,
Zope Enterprise Objects, extends the ZODB machinery to provide access
to objects over a network.  The name "Zope Enterprise Objects" is a
bit misleading; ZEO can be used to store Python objects and access
them in a distributed fashion without Zope ever entering the picture.
The combination of ZEO and ZODB is essentially a Python-specific
object database.

ZEO consists of about 1400 lines of Python code.  The code is
relatively small because it contains only code for a TCP/IP server,
and for a new type of Storage, \class{ClientStorage}.
\class{ClientStorage} doesn't use disk files at all; it simply
makes remote procedure calls to the server, which then passes them on
a regular \class{Storage} class such as \class{FileStorage}.  The
following diagram lays out the system:

XXX insert diagram here later

Any number of processes can create a \class{ClientStorage}
instance, and any number of threads in each process can be using that
instance.  \class{ClientStorage} aggressively caches objects
locally, so in order to avoid using stale data, the ZEO server sends
an invalidate message to all the connected \class{ClientStorage}
instances on every write operation.  The invalidate message contains
the object ID for each object that's been modified, letting the
\class{ClientStorage} instances delete the old data for the
given object from their caches.

This design decision has some consequences you should be aware of.
First, while ZEO isn't tied to Zope, it was first written for use with
Zope, which stores HTML, images, and program code in the database.  As
a result, reads from the database are \emph{far} more frequent than
writes, and ZEO is therefore better suited for read-intensive
applications.  If every \class{ClientStorage} is writing to the
database all the time, this will result in a storm of invalidate
messages being sent, and this might take up more processing time than
the actual database operations themselves.

On the other hand, for applications that have few writes in comparison
to the number of read accesses, this aggressive caching can be a major
win.  Consider a Slashdot-like discussion forum that divides the load
among several Web servers.  If news items and postings are represented
by objects and accessed through ZEO, then the most heavily accessed
objects -- the most recent or most popular postings -- will very
quickly wind up in the caches of the
\class{ClientStorage} instances on the front-end servers.  The
back-end ZEO server will do relatively little work, only being called
upon to return the occasional older posting that's requested, and to
send the occasional invalidate message when a new posting is added.
The ZEO server isn't going to be contacted for every single request,
so its workload will remain manageable.

\subsection{Installing ZEO}

This section covers how to install the ZEO package, and how to 
configure and run a ZEO Storage Server on a machine. 

\subsubsection{Requirements}

To run a ZEO server, you'll need Python 1.5.2 or 2.0, and the ZODB
packages from \url{http://www.amk.ca/files/zodb/}
have to be installed.  

\emph{Note for Python 1.5.2 users}: ZEO requires updated versions
of the \module{asyncore.py} and \module{asynchat.py} modules that are
included in 1.5.2's standard library.  Current versions of the ZODB
distribution install private versions of these modules, so you
shouldn't need to grab updated versions yourself.  (The symptom of
this problem is a traceback on attempting to run a ZEO client program:
the traceback is ``TypeError: too many arguments; expected 2, got 3''
around line 100 of \file{smac.py}.

\subsubsection{Installation}

Installing the ZEO package is easy.  Just run \code{python setup.py
install}.  This will install the ZEO/ package into your Python
installation, and copy various files into their proper locations:
\file{zeo.conf} will be put into \file{/usr/local/etc/}, a \file{zeo} startup
script will be put in \file{/etc/rc.d/init.d/}, and the \file{zeod}
daemon program will be placed in \file{/usr/local/bin}.

\subsection{Configuring and Running a ZEO Server}

Edit \code{/usr/local/etc/zeo.conf} appropriately for your desired
setup.  This configuration file controls the port on which ZEO will
listen for connections, the user and group IDs under which the server
will be executed, and the location of the concrete \class{Storage}
object that will be made network-accessible.
 
\subsection{Testing the ZEO Installation}

Once a ZEO server is up and running, using it is just like using ZODB
with a more conventional disk-based storage; no new programming
details are introduced by using a remote server.  The only difference
is that programs must create a \class{ClientStorage} instance instead
of a \class{FileStorage} instance.  From that point onward, ZODB-based
code is happily unaware that objects are being retrieved from a ZEO
server, and not from the local disk.

As an example, and to test whether ZEO is working correctly, try
running the following lines of code, which will connect to the server,
add some bits of data to the root of the ZODB, and commits the
transaction:

\begin{verbatim}
from ZEO import ClientStorage
from ZODB import DB

# Change next line to connect to your ZEO server
addr = ('kronos.example.com', 1975)
storage = ClientStorage.ClientStorage(addr)
db = DB(storage)
conn = db.open()
root = conn.root()

# Store some things in the root
root['list'] = ['a', 'b', 1.0, 3]
root['dict'] = {'a':1, 'b':4}

# Commit the transaction
get_transaction().commit()
\end{verbatim}

If this code runs properly, then your ZEO server is working correctly.

\subsection{ZEO Programming Notes}

XXX The Connection.sync() method and its necessity (if it works at all!) 

% That doesn't work.  I tested it.  sync() doesn't seem to get into
% the asyncore loop.  One of us should probably look into adding an
% API for this when we have some free time.  It would be a nice
% small project that would get into ZODB's guts.




\subsection{Sample Application: chatter.py}

For an example application, we'll build a little chat application.
What's interesting is that none of the application's code deals with
network programming at all; instead, an object will hold chat
messages, and be magically shared between all the clients through ZEO.
I won't present the complete script here; it's included in my ZODB
distribution, and you can download it from
\url{http://www.amk.ca/zodb/demos/}.  Only the interesting portions of
the code will be covered here.

The basic data structure is the \class{ChatSession} object,
which provides an \method{add_message()} method that adds a
message, and a \method{new_messages()} method that returns a list
of new messages that have accumulated since the last call to
\method{new_messages()}.  Internally, \class{ChatSession}
maintains a B-tree that uses the time as the key, and stores the
message as the corresponding value.

The constructor for \class{ChatSession} is pretty simple; it simply
creates an attribute containing a B-tree:

\begin{verbatim}
class ChatSession(Persistent):
    def __init__(self, name):
        self.name = name
        # Internal attribute: _messages holds all the chat messages.        
        self._messages = BTree.BTree()        
\end{verbatim}

\method{add_message()} has to add a message to the
\code{_messages} B-tree.  A complication is that it's possible
that some other client is trying to add a message at the same time;
when this happens, the client that commits first wins, and the second
client will get a \exception{ConflictError} exception when it tries to
commit.  For this application, \exception{ConflictError} isn't serious
but simply means that the operation has to be retried; other
applications might treat it as a fatal error.  The code uses
\code{try...except...else} inside a \code{while} loop,
breaking out of the loop when the commit works without raising an
exception.

\begin{verbatim}
    def add_message(self, message):
        """Add a message to the channel.
        message -- text of the message to be added
        """

        while 1:
            try:
                now = time.time()
                self._messages[now] = message
                get_transaction().commit()
            except ConflictError:
                # Conflict occurred; this process should pause and
                # wait for a little bit, then try again.
                time.sleep(.2)
                pass
            else:
                # No ConflictError exception raised, so break
                # out of the enclosing while loop.
                break
        # end while
\end{verbatim}

\method{new_messages()} introduces the use of \textit{volatile}
attributes.  Attributes of a persistent object that begin with
\code{_v_} are considered volatile and are never stored in the
database.  \method{new_messages()} needs to store the last time
the method was called, but if the time was stored as a regular
attribute, its value would be committed to the database and shared
with all the other clients.  \method{new_messages()} would then
return the new messages accumulated since any other client called
\method{new_messages()}, which isn't what we want.

\begin{verbatim}
    def new_messages(self):
        "Return new messages."

        # self._v_last_time is the time of the most recent message
        # returned to the user of this class. 
        if not hasattr(self, '_v_last_time'):
            self._v_last_time = 0

        new = []
        T = self._v_last_time

        for T2, message in self._messages.items():
            if T2 > T:
                new.append(message)
                self._v_last_time = T2

        return new
\end{verbatim}

This application is interesting because it uses ZEO to easily share a
data structure; ZEO and ZODB are being used for their networking
ability, not primarily for their data storage ability.  I can foresee
many interesting applications using ZEO in this way:

\begin{itemize}
  \item With a Tkinter front-end, and a cleverer, more scalable data
  structure, you could build a shared whiteboard using the same
  technique.

  \item A shared chessboard object would make writing a networked chess
  game easy.  

  \item You could create a Python class containing a CD's title and
  track information.  To make a CD database, a read-only ZEO server
  could be opened to the world, or an HTTP or XML-RPC interface could
  be written on top of the ZODB.

  \item A program like Quicken could use a ZODB on the local disk to
  store its data.  This avoids the need to write and maintain
  specialized I/O code that reads in your objects and writes them out;
  instead you can concentrate on the problem domain, writing objects
  that represent cheques, stock portfolios, or whatever.

\end{itemize}

